A family of accessibility measures derived from spatial interaction principles

### 2025-10-22 paper_plos/paper_plos_V3.qmd
V3 final edits AS

# Introduction

@@

In summary, this paper makes two contributions: (1) we review how spatial interaction modeling and accessibility share similar origins but have diverged in focus and interpretation. (2) we introduce a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population". These cases and variants align with common measures such as Hansen-type accessibility [@hansen1959], competitive accessibility measures such as the 2SFCA method [@shen1998; @luo2003], and market potential models [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].

@@

The remainder of this paper proceeds as follows. Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and @careyPrinciplesSocialScience1858 (Section 2), moving through early researchers like @ravensteinLawsMigration1885 to @stewartDemographicGravitationEvidence1948 who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in @hansen1959 became the dominant approach in planning practice (Section 4). Section 5 presents the entropy-based family of spatial interaction models in @wilson1971, explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates. Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations. Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied -- unconstrained, singly constrained, and doubly constrained cases -- demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices. We illustrate each measure with numerical examples that clarify the practical implications of different constraint assumptions. Section 8 concludes by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.

@@

# Newtonian's roots of human spatial interaction research

@@

# Early research on human spatial interaction: from Ravenstein (1885) to Stewart (1948)

@@

# Hansen's gravity-based accessibility to today {#grav-to-today}

@@

# Wilson's family of spatial interaction models
While accessibility research evolved in North America with Hansen [@hansen1959], a parallel development was taking place across the Atlantic with Alan G. Wilson. Wilson's ground breaking paper [@wilson1971] advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy. Additionally, this work was not focused on the 'potential' concept as accessibility had been. Wilson [@wilson1971] formalized the general spatial interaction model through the following equation:

@@

Wilson's work is notable for many reasons. Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation. This results in interpretable, balanced (given the system knowns are expressed in the constraints), and unit-consistent $ij$ flows. The balancing factors themselves have been subject to various interpretations -- as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @Wilson1971) and as rents [@morphetGravityModelCalibration2023] -- reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally. In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled. While Wilson's model produces results that are in units of flow tethered to the system of analysis, which facilitates such interpretations, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units.

@@

# Accessibility and spatial interaction modelling: two divergent research streams

@@

# A family of accessibility measures: from proportionality to equality

@@

## Toy example setup

@@

Any set of concepts representing population, opportunities, and their associated travel behaviour, whether representing the entire region uniformly (as will be demonstrated) or representing specific subgroups, can be substituted into our simple toy example. The purpose of the following simple example is to demonstrate the calculation and interpretation of the four accessibility measure variants.

## Unconstrained accessibility

@@

For example, @tbl-simple-example-unconstrained-accessibility shows $V^0_{i}$ under each decay function. Comparing across decay types is meaningless in absolute terms. For instance, the difference in zone 1 (edge of urban core)'s accessibility under $f_3$ vs $f_1$ is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units? These two values are a product of different impedance functions ( _physicians-minute$^{-3}$_ and _physicians-minute$^{-0.1}$_), making the comparison uninterpretable (and arguably incorrect). The fundamental uninterpretability of what is an _opportunity-weighted-travel-impedance_ unit remains.

@@

## Total constrained accessibility

@@

### Total constrained accessible opportunities: Hansen's accessibility with a total constraint
In the total constrained case, accessibility is expressed as a share of the total number of opportunities in the region $D$, allocated based on travel impedance. The total constrained accessibility from $i$ to $j$ takes the form:
$$
V^T_{ij} = \kappa_{ij}^T D
$$ {#eq-total-constrained-access}

@@

### Total constrained accessible population: Reilly's potential trade territories with a total constraint

@@

## Singly constrained accessibility

@@

### Singly constrained accessible opportunities: spatial availability

@@

### Singly constrained accessible population: market availability

@@

## Doubly constrained accessibility

@@

By contrast, the doubly constrained model assumes both the demand (population) and the supply (opportunity) are known and bounded, and allocates flows accordingly. This makes it less suitable for traditional accessibility analysis; namely because origin and destination masses often differ in kind and units. For instance, the number of people accessing an opportunity such as park may be known, but the capacity of each park is not. A doubly constrained approach may make conceptual sense if the "potential" should be contained only in the flows themselves, meaning the units of population and opportunities are comparable, have a one-to-one correspondence or are otherwise paired (e.g., job per worker, student per school-seat, or vaccine doses per person).

As mentioned, doubly constrained approach to accessibility adopts the structure of the production-attraction spatial interaction model, where both population and opportunity totals are fixed. Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality-2):

@@

```{r}
#| label: tbl-small-system-land-use-doubly constrained-case
#| tbl-cap: "Modified simple system with three zones reflecting matched population and opportunities. Population is in 10,000 persons and opportunities in 10,000 of physician-capacity."

@@

As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints -- i.e., $V^D_i = O_i$ so the physician-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`. Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values. $V^D_{ij}$ values represent the number of opportunities from zone $j$ allocated to populations in zone $i$, shaped by both mass and travel impedance.

To illustrate this, consider the results for Zone 2 (recall: a suburban zone -- a high-population, low-opportunity, relatively remote). As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal flow (i.e., $V^D_{22}$) declines as travel impedance decay decreases -- from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)|> round(digits = 3)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3) |> round(digits = 3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).

@@

Recall, accessibility is conventionally presented as a zonal value not a flow. However, in the doubly constrained case, since we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, $V_i^D$ is simply the number of opportunities that matches our known population at $i$. So following the logic of the family of accessibility measures, in the doubly constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity. Furthermore, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $\kappa_{ij}^D$ and $\hat \kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.

@@

And as mentioned, these units are often misaligned -- for instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person. When they do align -- such as people to physician capacity -- we are modeling realised access flows based on known quantities of _those that interact_ and the _interacted_. In such cases, the traditional accessibility question can be seen to be already answered by the known information (i.e., how many opportunities can be reached by a zone? Answer: the number of people at that zone). For this reason, we do not foresee the doubly constrained measure being widely used in accessibility analysis as the literature has largely focused on questions of 'potential' access, not on predicting flows of realised access.

# Conclusions

@@

Lastly, the doubly constrained case is introduced. In this case, the sums must equal both the regional total and ensure that no zone allocates more opportunities than it has available. Specifically, accessibility values for each $i$-$j$ pair must be a proportion of the zonal opportunity and population values simultaneously. For example, the accessibility at Zone 1 must equal the sum of opportunities from zones 1, 2, and 3, as well as the sum of the population at Zone 1. Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair values should be of interest. In this sense, the research question should be concerned with 'access' (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential spatial interaction (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach out of a regional total and/or zonal-allocation).

Overall, and building on Wilson's [@wilson1971] foundational work, this paper proposes a unified framework for analyzing gravity-based accessibility. By reintroducing Wilson's proportionality constants, the family of constrained accessibility measures restores measurement units to accessibility estimates. This enhancement provides a more interpretable, consistent, and theoretically grounded basis for accessibility analysis, which could help advance the adoption of accessibility-oriented planning.

That said, there have been more recent developments in the accessibility literature, including novel person-based approaches that are time-sensitive [@yang2024evaluating; @braga2023evaluating], behavioural [@kar2024inclusive; @lu2014effects] or utility-based [@guzman2023much; @ben1985discrete], that are not based on top-down approaches such as Wilson's model formalized through the entropy-maximisation procedure. While gravity-based measures of accessibility dominate the applied literature, future work could further explore how these ideas of proportionality constants and balanced units could also help inform these and other new modeling approaches.

>>>>

# 1. Introduction

@@

This paper makes two contributions. First, we review how spatial interaction modeling and accessibility share similar origins but have diverged in focus and interpretation. Second, we introduce a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population". These cases and variants align with common measures such as Hansen-type accessibility [@hansen1959], competitive accessibility measures such as the 2SFCA method [@shen1998; @luo2003], and market potential models [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].

@@

The remainder of this paper proceeds as follows. Sections 2 through 4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and Carey (1858) [@careyPrinciplesSocialScience1858] (Section 2), moving through early researchers like Ravenstein (1885) [@ravensteinLawsMigration1885] to Stewart (1948) [@stewartDemographicGravitationEvidence1948] who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in Hansen (1959) [@hansen1959] became the dominant approach in planning practice (Section 4). Section 5 presents the entropy-based family of spatial interaction models in Wilson (1971) [@wilson1971], explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates. Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations. Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied -- unconstrained, singly constrained, and doubly constrained cases -- demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices. We illustrate each measure with numerical examples that clarify the practical implications of different constraint assumptions. Section 8 concludes by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.

@@

# 2. Newtonian's roots of human spatial interaction research

@@

# 3. Early research on human spatial interaction: from Ravenstein (1885) to Stewart (1948)

@@

# 4. Hansen's gravity-based accessibility to today {#grav-to-today}

@@

# 5. Wilson's family of spatial interaction models
While accessibility research evolved in North America with Hansen [@hansen1959], a parallel development was taking place across the Atlantic with Alan G. Wilson. Wilson's ground breaking paper [@wilson1971] advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy. Additionally, this work was not focused on the 'potential' concept as accessibility had been. Wilson [@wilson1971] formalized the general spatial interaction model through the following equation:

@@

Wilson's work is notable for many reasons. Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation. This results in interpretable, balanced (given the system knowns are expressed in the constraints), and unit-consistent $ij$ flows. The balancing factors themselves have been subject to various interpretations -- as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in Wilson [@wilson1971]) and as rents [@morphetGravityModelCalibration2023] -- reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally. In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled. While Wilson's model produces results that are in units of flow tethered to the system of analysis, which facilitates such interpretations, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units.

@@

# 6. Accessibility and spatial interaction modelling: two divergent research streams

@@

# 7. A family of accessibility measures: from proportionality to equality

@@

## 7.1. Toy example setup

@@

Any set of concepts representing population, opportunities, and their associated travel behaviour, whether representing the entire region uniformly (as will be demonstrated) or representing specific subgroups, can be substituted into our simple toy example. The purpose of the following simple example is to demonstrate the calculation and interpretation of the four accessibility measure variants. As well, the family of measures proposed will be available in the [{accessibility} R package](https://ipeagit.github.io/accessibility/) for convenient use in future studies.

## 7.2. Unconstrained accessibility

@@

For example, @tbl-simple-example-unconstrained-accessibility shows $V^0_{i}$ under each decay function. Comparing across decay types is meaningless in absolute terms. For instance, the difference in zone 1 (edge of urban core)'s accessibility under $f_3$ vs $f_1$ is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units? These two values are a product of different impedance functions _physicians-minute$^{-0.1}$_ and _physicians-minute$^{-3}$_), making the direct comparison uninterpretable (and arguably incorrect). The fundamental uninterpretability of what is an _opportunity-weighted-travel-impedance_ unit remains.

@@

## 7.3. Total constrained accessibility

@@

### 7.3.1. Total constrained accessible opportunities: Hansen's accessibility with a total constraint
In the total constrained case, accessibility is expressed as a share of the total number of opportunities in the region $D$, allocated based on travel impedance. The total constrained accessibility from $i$ to $j$ takes the form:
$$
V^T_{ij} = \kappa_{ij}^T D
$$ {#eq-total-constrained-access}

@@

### 7.3.2. Total constrained accessible population: Reilly's potential trade territories with a total constraint

@@

## 7.4. Singly constrained accessibility

@@

### 7.4.1. Singly constrained accessible opportunities: spatial availability

@@

### 7.4.2. Singly constrained accessible population: market availability

@@

## 7.5. Doubly constrained accessibility

@@

By contrast, the doubly constrained model assumes both the demand (population) and the supply (opportunity) are known and bounded, and allocates flows accordingly. This makes it less suitable for traditional accessibility analysis; namely because origin and destination masses often differ in kind and units. For instance, the number of people accessing an opportunity such as park may be known, but the capacity of each park is not. A doubly constrained approach may make conceptual sense if the "potential" should be contained only in the flows themselves, meaning the units of population and opportunities are comparable, have a one-to-one correspondence or are otherwise paired (e.g., job per worker, student per school-seat, or vaccine doses per person). Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality-2):

@@

```{r}
#| label: tbl-small-system-land-use-doubly-constrained-case
#| tbl-cap: "Modified simple system with three zones reflecting matched population and opportunities. Population is in 10,000 persons and opportunities in 10,000 of physician-capacity."

@@

As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints i.e., $V^D_i = O_i$ hence the physician-capacity accessible for Zones 1, 2, and 3 are `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`. The usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values; the number of opportunities from zone $j$ allocated to populations in zone $i$ shaped by both mass and travel impedance.

To illustrate this concept, consider the results for Zone 2 (recall: a suburban type zone, with a higher population, lower amount of opportunities, and relatively remote). As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal flow (i.e., $V^D_{22}$) declines as travel impedance decay decreases -- from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)|> round(digits = 3)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3) |> round(digits = 3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).

@@

Accessibility is conventionally presented as a zonal value not a flow. However, in the doubly constrained case, since we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, $V_i^D$ is simply the number of opportunities that matches our known population at $i$. Following the logic of the family of accessibility measures, in the doubly constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity. Furthermore, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $\kappa_{ij}^D$ and $\hat \kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.

@@

And as mentioned, these units are often misaligned. For instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person. When they do align, such as people to physician capacity, we would be modeling realised access flows based on known quantities of _those that interact_ and the _interacted_. In such cases, the traditional accessibility question can be seen to be already answered by the known information (i.e., how many opportunities can be reached by a zone? Answer: the number of people at that zone). For this reason, we do not foresee the doubly constrained measure being widely used in accessibility analysis as the literature has largely focused on questions of 'potential' not on predicting flows of realised access.

# 8. Conclusions

@@

Lastly, the doubly constrained case is introduced. In this case, the sums must equal both the regional total and ensure that no zone allocates more opportunities than it has available. Specifically, accessibility values for each $i$-$j$ pair must be a proportion of the zonal opportunity and population values simultaneously. For example, the accessibility at Zone 1 must equal the sum of opportunities from Zones 1, 2, and 3, as well as the sum of the population at Zone 1. Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair value should be of research interest.

Building on Wilson's [@wilson1971] foundational work, this paper proposes a unified framework for analyzing gravity-based accessibility. By reintroducing Wilson's proportionality constants, the family of constrained accessibility measures restores measurement units to accessibility estimates. This enhancement provides a more interpretable, consistent, and theoretically grounded basis for accessibility analysis, which could help advance the adoption of accessibility-oriented planning.

While this work exclusively focused on top-down gravity-based accessibility measures, there have been recent developments in the accessibility literature that include person-based approaches that are time-sensitive [@yang2024evaluating; @braga2023evaluating], behavioural [@kar2024inclusive; @lu2014effects] or utility-based [@guzman2023much; @ben1985discrete]. While gravity-based measures of accessibility still dominate the applied literature, future work could further explore how these ideas of proportionality constants and balanced units could also help inform these and other new modeling approaches.
